
---
title: "PROJECT 01: Spatial Distribution of Fire Incidents in D.C., Maryland, and Virginia in 2018"
author: |
  Jeongwon Yoo  
  Simbanegavi Simbarashe  
  Nithin Ravindra Reddy  
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r setup, include=FALSE}
# Set CRAN mirror
options(repos = c(CRAN = "https://cloud.r-project.org"))

# Install required packages if they are not already installed
required_packages <- c("sf", "ggplot2", "viridis", "dplyr", "lubridate")
new_packages <- required_packages[!(required_packages %in% installed.packages()[, "Package"])]
if (length(new_packages)) {
  install.packages(new_packages)
}

# Load libraries
library(sf)
library(ggplot2)
library(viridis)
library(dplyr)
library(lubridate)
library(tigris)
library(readr)
library(tidyr)
library(corrplot)
library(spdep)
library(spatialreg)

```


```{r}
# Set global chunk options
knitr::opts_chunk$set(echo = TRUE, results = "markup", warning = FALSE, message = FALSE)

# Set options for number display format
options(scientific = TRUE, digits = 3)
```
#  CHAPTER 1: Introduction

## 1.0 Overview

 
## 1.1 Objectives



## 1.2 Research Questions


## 1.3 Significance of the Study



## 1.4 Weakness of the Study

# 1.4.1 Data Limitations


# 1.4.2 Methodological Weaknesses



# 1.4.3 Scope and Generalizability



#  CHAPTER 2: Methodology
## 2.0 Dataset




## 2.1 Software

The following software was used for this project:

R and R Markdown

GitHub (GitHub Repository)

Microsoft Excel

## 2.2 Project Study Area
The study was conducted in 3 states namely DC,VA and MD

```{r load_and_display_study_area}

# Specify the path to your study area shapefile
shapefile_path <- "C:/Users/SIMBAH/Desktop/Intro to Data science/Project/DC_MD_VA Boundaries/Project_Area.shp"

# Load the shapefile
Study_Area <- st_read(shapefile_path)

# Check the CRS
print(st_crs(Study_Area))

# Transform to NAD83 / UTM Zone 18N (EPSG:26918)
Study_Area <- st_transform(Study_Area, crs = 26918)

# View the structure of the shapefile data
#str(Study_Area)

# Plot the data with labels
library(ggplot2)
library(viridis)

ggplot(data = Study_Area) +
  geom_sf(aes(fill = NAME_1), color = "Green") +  # Fill by NAME_1 and add a border
  geom_sf_text(aes(label = NAME_1), size = 3, check_overlap = TRUE, color = "black", fontface = "bold") +  # Add labels in bold black
  scale_fill_viridis_d(option = "C", name = "Key") +  # Update legend title to "Key"
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),  # Center the title
    legend.position = "right"  # Position the legend on the right
  ) +
  ggtitle("Project Study Area Map")
```

# CHAPTER 2: Analysis and Results

# 3.1 Objective 1: Categorize Types of Fire Incidents and Assess Causes



## 3.1.1 Read dataset



## 3.1.2 Quick overview


## 3.1.3 summary 


## 3.1.4 CATEGORIZATION: Fire Incident Types and Causes



## 3.1.5 Bar Chart


## 3.1.6 Frequency count of causes


## 3.1.7 Segregation based on states



## 3.1.8 Bar chart on fire incidences in States.


## 3.1.9 Classification of urban and rural counties in DC,MD and VA.


## 3.1.10 Frequency table of causes by area type



## 3.1.11 Distibution of fire cause by area type

# 3.2 Objective 2

## 3.2.1 EXPLORATORY DATA ANALYSIS


## 3.2.5 Fire Incidents by Cause



## 3.2.8 Fire Incidences by Ownership



# 3.2.9 Average fire size by cause




##  Chapter 4: Conclusion and Recommendations

## 4.1 Conclusion


# 4.2 Recommendations



# REFERENCES

Fire Loss in the United States 2020

National Emergency Response Information System (NERIS) 2019

National Fire Incident Reporting System (NFIRS) 2018

National Fire Protection Association (NFPA) 2019

National Inter-agency Fire Center (NIFC) 2020

U.S. Fire Administration (USFA) 2018

USFA Topical Fire Reports 2018

NFPA Survey Estimates 2019

